\thetitle

### 2020-07-07 writing/rational.tex
ogranizing section 5

\section{How to Simulate the Rational Programmer} \label{sec:rational}

Consider a rational programmer  that tries to debug a faulty program in
natural Typed Racket. If the Typed Racket
type checker detects the problem, the programmer knows where
to look for the bug; in the component that contains the type specification
the type checker complains about. However, if
the type checker blesses the program, 
the programmer relies on  runtime type checks  to  signal the impedance mismatch between the
program's behavior and its types. If these checks do raise an error that
blames a component in the program, 
the rational programmer ``follows'' blame to
locate where to add a type specification next.  Hopefully, the extra
types are sufficient to turn the run time error into a static one. Otherwise,
the rational programmer starts over with the latest version of the program. 
In other words, at each iteration the
faulty program migrates to a version with more type specifications until
it becomes ill-typed.

Therefore, a concise way to model rational programmers is with the
migratory paths they follow through the \emph{configuration space} of a
program. This space consists of variants of the program, dubbed
\emph{configurations}, that differ in their type specifications. In the 
setting of Typed Racket's migratory typing,\footnote{The
method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.} each
configuration differs from another in the components that have type
specifications. Hence, formally, we can represent a program
$\system$ as a set of components $\set{\component}$ and each
configuration $\conf$  as a subset of $\set{\component}$ that contains
exactly the components of  $\system$ that are typed in $\conf$. Naturally
the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$ elements.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations.

With these definitions, the actions of the rational programmer map to an ascending chain in
$\lattice{\system}$, dubbed a \emph{trail}. 
The starting configuration $\conf_0$ of a trail,  dubbed the \emph{root of the
trail}, is the configuration that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that 
the untyped component  that $\conf_i$ blames under the natural semantics, dubbed $\blame{\system, \conf_i}$,
 becomes typed in $\conf_{i+1}$.\footnote{Blame can result either directly
 from a run time type check or, in the absence of blame, from an
 exception that we track back to the component that causes it with the
 help of the stack trace.} Putting these pieces together:

@@

\subsection{How to Generalize the Rational Programmer Beyond Natural}

Adapting the simulation of rational programmer to transient
and erasure requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame. At the other end of the blame spectrum, erasure
blames no component and relies on the exceptions of the underlying
language.

Hence, for an erasure rational programmer only the exception and random 
rational programmers are meaningful. In contrast, for transient, we 
need two transient-specific modes for the rational
programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn  trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ denotes the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}

@@

In the case of erasure, instead of blame, we can examine
the value of runtime exceptions for the rational programmer. Thus, in the
same vain as for transient and natural we
ask whether the configuration lattice of a given program 
contains only successful erasure exception trail and at least one failing
erasure random trail. Of course, we can formulate the same questions about 
 exceptions for natural and transient and query the value of
exceptions for these semantics too.

Overall our experiment examines six questions for each benchmark: one for the value of natural
blame, two for the value of transient blame and one for the value of
exceptions in each of natural, transient and erasure.
Table~\ref{fig:experiment-outline} summarizes how each question relates to different
trails/modes of the rational programmer. For example 
experimental question $Q_1$ asks whether blame adds value to transient and 
our experiment uses the natural blame, exception and random trails to
answer it.

\begin{figure}
\center
{\begin{tabular}[c]{c|c|c|c}
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure}  \\
\hline 
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
{\bf Random}            &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes. $Q_1$:
  Does natural blame add value to natural? $Q_2$: Does transient all blame
  add value to transient?  $Q_3$: Does transient firstborn blame add value
  to transient?  $Q_4$: Do exceptions add value to natural?  $Q_5$: Do
  exceptions add value to transient? $Q_6$: Do exceptions add value to
  erasure? } 
  \label{fig:experiment-outline}
\end{figure}

\noindent
{\bf Programmer Effort.} In addition to the value of blame, trails permit
us to examine the effort a rational programmer spends to locate a bug. 
For instance, the length of a successful natural blame trail tells us how
many components the programmer has to type before reaching the faulty
component. This idea also applies to all other modes of the rational
programmer. The most interesting case is that for transient all blame trails. 
In detail, given a successful transient all blame trail 
$\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,  
the effort of the rational programmer is the 
number of distinct configurations in the all blame trails with root $\conf_n$ and 
length $n$ in $\lattice{\system}$. Our experiment utilizes this notion of
effort to analyze burden on the rational programmer for finding a
bug when the answer to one of the experimental questions from above is
positive for a benchmark.

>>>>

\section{How to Simulate the Rational Programmer}

Code migration is the raison d'\^etre of Typed Racket, which is the community
uses ``migratory typing'' instead of gradual typing. Roughly speaking a
programmer uses a gradually typed language to incrementally add types to a code
base. From this perspective, a code base consists of some number of components,
say modules, and a programmer converts a component at a time from its untyped
state to a typed one.

Based on this idea, a metric for migratory paths is a concise way to measure a
rational programmer. The components of the program create a \emph{configuration
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{configurations}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each configuration differs from another in the subset
of typed components. Formally, a program $\system$ is a set of components
$\set{\component}$, and a configuration $\conf$ is a subset of
$\set{\component}$ . These configurations of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
and the top element is $\system$ itself. In between are all the mixed-type
configurations.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The starting
configuration $\conf_0$ of a trail---also referred to the \emph{root} (of the
trail)---is the configuration that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that the untyped component that $\conf_i$ blames under the
various semantics, dubbed $\blame{\system, \conf_i}$, becomes typed in
$\conf_{i+1}$.\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to the
component that causes it with the help of the stack trace.}

The next three subsections specialize the idea to the various semantics of
gradual typing. Once this machinery is in place, we use the fourth subsection to
state the precise experimental research questions that correspond to the
high-level question on the first page. The final subsection explains how the
rational programmer's effort is measured.

%% -----------------------------------------------------------------------------
\subsection{Modes of Natural Type Programming} \label{sub:natural}

Putting these pieces together:

@@

%% -----------------------------------------------------------------------------
\subsection{Modes of Optional Type Programming} \label{sub:transient}

The case of erasure semantics is analogous to the third mode of natural.  Since
gradually typed languages with erasure semantics do not come with blame
assignment, a rational programmer can only hope that the underlying safety
checks and their stack traces are helpful.  Thus, in the same veub as for
natural we ask whether the configuration lattice of a given program contains
only successful erasure exception trail and at least one failing erasure random
trail.

%% -----------------------------------------------------------------------------
\subsection{Modes of Transient Type Programming} \label{sub:transient}

Adapting the simulation of rational programmer to transient
requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame.

For transient, we 
need two transient-specific modes for the rational
programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn  trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ denotes the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}

@@

%% -----------------------------------------------------------------------------
\subsection{Questions}

In sum, our experiment examines four questions for each benchmark:
\begin{itemize}
\item[$Q_1$] Does natural blame add value to natural?

\item[$Q_2$] Does transient all blame add value to transient?

\item[$Q_3$] Does transient firstborn blame add value to transient?

\item[$Q_*$] Do exceptions add value? For any of the systems? 
\end{itemize}

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame adds value to transient and our experiment
uses the natural blame, exception and random trails to answer it.

\begin{figure}
\center
{\begin{tabular}[ht]{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

%% -----------------------------------------------------------------------------

\subsection{Programmer Effort}

In addition to the value of blame, trails permit us to examine the effort a
rational programmer spends to locate a bug.  For instance, the length of a
successful natural blame trail tells us how many components the programmer has
to type before reaching the faulty component. This idea also applies to all
other modes of the rational programmer. The most interesting case is that for
transient all blame trails.  In detail, given a successful transient all blame
trail $\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,
the effort of the rational programmer is the number of distinct configurations
in the all blame trails with root $\conf_n$ and length $n$ in
$\lattice{\system}$. Our experiment utilizes this notion of effort to analyze
burden on the rational programmer for finding a bug when the answer to one of
the experimental questions from above is positive for a benchmark.
